---
title: "Lab 8 Machine Learning 1"
author: 'Thao Hoang PID: A15449860'
date: "10/21/2021"
output: pdf_document
---

First up is clustering method

#Kmeans clustering 

The function in base R to do Kmeans clustering is called `kmeans()`.

First make up some data where we know what the answer should be:

```{r}
tmp <- c(rnorm(30,-3), rnorm(30,3))
x<- cbind(x=tmp, y=rev(tmp))
plot(x)
```

> Q. Can we use kmeans() to cluster this data setting k 2 and nstart to 20?

```{r}
 km <- kmeans(x, centers=2, nstart=20 )
km 
```

> Q. how many points are in each cluster? 

```{r}
km$size
```
> Q. What 'components' of your result object details cluster assignmnet/membership?

```{r}
km$cluster
```

>Q. Can you find the cluster center?

```{r}
km$centers
```

> Q. plot x colored by the kmeans cluster assignment and add cluster centers as blue points

```{r}
plot(x, col=km$cluster)
points(km$centers, col="blue", pch=15, cex=2)
       
```

# hclust (hierarchial clustering)

A big limittation with kmeans is that we have to tell it K (the number of clusters we want
)
Analyze this same data with hclust 

demonstrate the use of dist(), hclust(), plot(), and cutree() functions to do the clustering, 
Generate dendrograms and return cluster assignment / membership vector

```{r}
hc<- hclust(dist(x))
hc
```

There is a plot method for hclust result objects, let's see

```{r}
plot(hc)

```

to get our cluster membership vector we have to do a wee bit more work. We have to "cut" the tree where we think it makes sense. For this we use the 'cutree()' function.

```{r}
cutree(hc, h=6)
```

you can also call cutree() setting k= the number of groups/clusters you want.

```{r}
groups <- cutree(hc, k=2)
```

make our result plot

```{r}
plot(x, col=groups)
```

Read data on food in the UK
```{r}
url <- "https://tinyurl.com/UK-foods"
x <- read.csv(url)
```

> Q1. How many rows and columns are in your new data frame named x? What R functions could you use to answer this questions?

```{r}
dim(x)
```

Preview the first 6 rows
```{r}
View(x)
```

This should be a 17 X 4, has an extra column that we need to fix
```{r}
# Note how the minus indexing works
rownames(x) <- x[,1]
x <- x[,-1]
head(x)
#dangerous because it overrides your data if you keep running it
```

```{r}
url <- "https://tinyurl.com/UK-foods"
x <- read.csv(url, row.names=1)
head(x)
```
checking dim() again
```{r}
dim(x)
```
> Q2. Which approach to solving the ‘row-names problem’ mentioned above do you prefer and why? Is one approach more robust than another under certain circumstances?

```{r}

url <- "https://tinyurl.com/UK-foods"
x <- read.csv(url, row.names=1)
head(x) 
```
 

is better because the other one, if you keep clicking run, it will continue deleting a column from the data resulting in all data being deleted. 

>Spotting major differences and trends

```{r}
barplot(as.matrix(x), beside=T, col=rainbow(nrow(x)))
```

>Q3: Changing what optional argument in the above barplot() function results in the following plot?

```{r}
barplot(as.matrix(x), beside=F, col=rainbow(nrow(x)))
```

>Q5: Generating all pairwise plots may help somewhat. Can you make sense of the following code and resulting figure? What does it mean if a given point lies on the diagonal for a given plot?

```{r}
pairs(x, col=rainbow(10), pch=16)
```
> Q6. What is the main differences between N. Ireland and the other countries of the UK in terms of this data-set?
  
  The main difference between N.Ireland and the other countries of the UK seemingly is that they consume more of fesh potatoes than others. 




PCA to the rescue 
Main base R PCA function is called `prcomp()`
We will need to transpose the data. 
```{r}
pca <- prcomp( t(x) )
summary(pca)
```
```{r}
attributes(pca)
```


> Q7. Complete the code below to generate a plot of PC1 vs PC2. The second line adds text labels over the data points.

```{r}
# Plot PC1 vs PC2
plot(pca$x[,1], pca$x[,2], xlab="PC1", ylab="PC2", xlim=c(-270,500))
text(pca$x[,1], pca$x[,2], colnames(x))
```

>Q8. Customize your plot so that the colors of the country names match the colors in our UK and Ireland map and table at start of this document.

```{r}
country_cols <- c("blue","red", "green", "purple")
plot(pca$x[,1], pca$x[,2], xlab="PC1", ylab="PC2", xlim=c(-270,500))
text(pca$x[,1], pca$x[,2], colnames(x), col=country_cols)
```

calculate how much variation in the original data each PC accounts for

```{r}
v <- round( pca$sdev^2/sum(pca$sdev^2) * 100 )
v
## or the second row here...
z <- summary(pca)
z$importance
```

plot of the variances 
```{r}
barplot(v, xlab="Principal Component", ylab="Percent Variation")
```

Digging deeper (variable loadings)

```{r}
## Lets focus on PC1 as it accounts for > 90% of variance 
par(mar=c(10, 3, 0.35, 0))
barplot( pca$rotation[,1], las=2 )
```

>Q9: Generate a similar ‘loadings plot’ for PC2. What two food groups feature prominantely and what does PC2 maninly tell us about?

Two food groups that are featured prominently are "fresh potatoes" and "other veg. PC2 mainly tell us about of the 29% sample variance of the 17 dimensional variance that has not been covered by PC1. 

```{r}
par(mar=c(10, 3, 0.35, 0))
barplot( pca$rotation[,2], las=2 )
```


Biplots
Another way to view UK food data together with PCA plot.
```{r}
## The inbuilt biplot() can be useful for small datasets 
biplot(pca)
```

PCA of RNA seq data
```{r}
url2 <- "https://tinyurl.com/expression-CSV"
rna.data <- read.csv(url2, row.names=1)
head(rna.data)
```

>Q10: How many genes and samples are in this data set?

```{r}
dim(rna.data)
```
There are 100 genes and 10 samples in this data set.


Lets do PCA and plot the results:

```{r}
## Again we have to take the transpose of our data 
pca <- prcomp(t(rna.data), scale=TRUE)
 
## Simple un polished plot of pc1 and pc2
plot(pca$x[,1], pca$x[,2], xlab="PC1", ylab="PC2")
```
```{r}
summary(pca)
```

Bar plot summary 
```{r}
plot(pca, main="Quick scree plot")
```

calculate how much variation in the original data each PC accounts for:
```{r}
## Variance captured per PC 
pca.var <- pca$sdev^2

## Percent variance is often more informative to look at 
pca.var.per <- round(pca.var/sum(pca.var)*100, 1)
pca.var.per
```

We can use this to generate our own scree-plot like this
```{r}
barplot(pca.var.per, main="Scree Plot", 
        names.arg = paste0("PC", 1:10),
        xlab="Principal Component", ylab="Percent Variation")
```

Now lets make our main PCA plot a bit more attractive and useful…
```{r}
## A vector of colors for wt and ko samples
colvec <- colnames(rna.data)
colvec[grep("wt", colvec)] <- "red"
colvec[grep("ko", colvec)] <- "blue"

plot(pca$x[,1], pca$x[,2], col=colvec, pch=16,
     xlab=paste0("PC1 (", pca.var.per[1], "%)"),
     ylab=paste0("PC2 (", pca.var.per[2], "%)"))

text(pca$x[,1], pca$x[,2], labels = colnames(rna.data), pos=c(rep(4,5), rep(2,5)))
```

Using ggplot

```{r}
library(ggplot2)

df <- as.data.frame(pca$x)

# Our first basic plot
ggplot(df) + 
  aes(PC1, PC2) + 
  geom_point()

```

Adding a condition specific color and perhaps sample label aesthetics for wild-type and knock-out samples 

```{r}
# Add a 'wt' and 'ko' "condition" column
df$samples <- colnames(rna.data) 
df$condition <- substr(colnames(rna.data),1,2)

p <- ggplot(df) + 
        aes(PC1, PC2, label=samples, col=condition) + 
        geom_label(show.legend = FALSE)
p

```

And finally add some spit and polish
```{r}
p + labs(title="PCA of RNASeq Data",
       subtitle = "PC1 clealy seperates wild-type from knock-out samples",
       x=paste0("PC1 (", pca.var.per[1], "%)"),
       y=paste0("PC2 (", pca.var.per[2], "%)"),
       caption="BIMM143 example data") +
     theme_bw()
```

Gene loadings
```{r}
loading_scores <- pca$rotation[,1]

## Find the top 10 measurements (genes) that contribute
## most to PC1 in either direction (+ or -)
gene_scores <- abs(loading_scores) 
gene_score_ranked <- sort(gene_scores, decreasing=TRUE)

## show the names of the top 10 genes
top_10_genes <- names(gene_score_ranked[1:10])
top_10_genes 
```








