---
title: "Lab 9 mini project"
author: 'Thao Hoang PID: A15449860'
date: "10/30/2021"
output: pdf_document
---

Reading the data from WisconsinCancer.csv 
```{r}
fna.data <- "WisconsinCancer.csv"
wisc.df <- read.csv(fna.data, row.names=1)
wisc.df
```
```{r}
# We can use -1 here to remove the first column
wisc.data <- wisc.df[,-1]
```
```{r}
# Create diagnosis vector for later 
diagnosis <- as.numeric(wisc.df$diagnosis == "M")
```

>Q1. How many observations are in this dataset?
There are 569 observations

```{r}
dim(wisc.df)
```


>Q2. How many of the observations have a malignant diagnosis?
There are 212 malignant diagnosis.

```{r}
sum(diagnosis)
```


>Q3. How many variables/features in the data are suffixed with _mean?

```{r}
length(grep("_mean", colnames(wisc.data)))
```


`Performing PCA`
Perform principal component analysis (PCA) on wisc.data.
```{r}
# Check column means and standard deviations
colMeans(wisc.data)

apply(wisc.data,2,sd)
```
Execute PCA with the prcomp() function on the wisc.data, scaling if appropriate, and assign the output model to wisc.pr.
```{r}
# Perform PCA on wisc.data by completing the following code
wisc.pr <- prcomp(wisc.data, scale = TRUE) 
# Look at summary of results
summary(wisc.pr)
```

>Q4. From your results, what proportion of the original variance is captured by the first principal components (PC1)?

0.4427

>Q5. How many principal components (PCs) are required to describe at least 70% of the original variance in the data?

3 PCs are required for at least 70% of the original variance, PC1,PC2, and PC3. 

>Q6. How many principal components (PCs) are required to describe at least 90% of the original variance in the data?

7 PCs are required for at least 90% of the original variance. PC1-PC7.


`Interpreting PCA results` 
>Q7. What stands out to you about this plot? Is it easy or difficult to understand? Why?
This is a hot mess of a plot and we will need to generate our own plots to make sense of this PCA result.

```{r}
biplot(wisc.pr)
```
Using scatter plot 
```{r}
# Scatter plot observations by components 1 and 2
plot( wisc.pr$x[,1:2], col = diagnosis +1, 
     xlab = "PC1", ylab = "PC2")
```

>Q8. Generate a similar plot for principal components 1 and 3. What do you notice about these plots?

```{r}
plot(wisc.pr$x[,1:3], col = diagnosis +1, 
     xlab = "PC1", ylab = "PC3")
```

Using the ggplot2 package to make a more fancy figure of these results.

```{r}
# Create a data.frame for ggplot
df <- as.data.frame(wisc.pr$x)
df$diagnosis <- diagnosis

# Load the ggplot2 package
library(ggplot2)

# Make a scatter plot colored by diagnosis 
# diagnosis 1=M, 0 = B
ggplot(df) + 
  aes(PC1, PC2, color= diagnosis) + 
  geom_point()
```

`Variance explained`
```{r}
# Calculate variance of each component
pr.var <- wisc.pr$sdev^2
head(pr.var)
```
```{r}
# Variance explained by each principal component: pve
pve <- pr.var / sum(pr.var)

# Plot variance explained for each principal component
plot(pve, xlab = "Principal Component", 
     ylab = "Proportion of Variance Explained", 
     ylim = c(0, 1), type = "o")
```

```{r}
# Alternative scree plot of the same data, note data driven y-axis
barplot(pve, ylab = "Precent of Variance Explained",
     names.arg=paste0("PC",1:length(pve)), las=2, axes = FALSE)
axis(2, at=pve, labels=round(pve,2)*100 )
```

>Q9. For the first principal component, what is the component of the loading vector (i.e. wisc.pr$rotation[,1]) for the feature concave.points_mean?

```{r}
wisc.pr$rotation[,(colnames(wisc.data) == "concave.points_mean")]
```

>Q10. What is the minimum number of principal components required to explain 80% of the variance of the data?

5 PCs is required to explain 80% of the variance. PC1-PC5 will give 84.731%


`Hierarchical clustering`
The goal of this section is to do hierarchical clustering of the original data

First scale the wisc.data data and assign the result to data.scaled.
```{r}
# Scale the wisc.data data using the "scale()" function
data.scaled <- scale(wisc.data)

```
Calculate the (Euclidean) distances between all pairs of observations in the new scaled dataset and assign the result to data.dist.
```{r}
data.dist <- dist(data.scaled)
```
Create a hierarchical clustering model using complete linkage. Manually specify the method argument to hclust() and assign the results to wisc.hclust.
```{r}
wisc.hclust <- hclust(data.dist, method = "complete")
```

>Q11. Using the plot() and abline() functions, what is the height at which the clustering model has 4 clusters?

```{r}
plot(wisc.hclust)
abline(h=19, col="red", lty=2)
```
Use cutree() to cut the tree so that it has 4 clusters. Assign the output to the variable wisc.hclust.clusters.
```{r}
wisc.hclust.clusters <- cutree(wisc.hclust, k = 4)
```

We can use the table() function to compare the cluster membership to the actual diagnoses.
```{r}
table(wisc.hclust.clusters, diagnosis)
```

>Q12. Can you find a better cluster vs diagnoses match by cutting into a different number of clusters between 2 and 10?

No, K=4 already covered the majority of the clustervs diagnoses matches. 

```{r}
wisc.hclust.clusters <- cutree(wisc.hclust, k = 7)
table(wisc.hclust.clusters, diagnosis)
```

>Q13. Which method gives your favorite results for the same data.dist dataset? Explain your reasoning.

The complete method since it gives the cleanest data.

```{r}
wisc.hclust <- hclust(data.dist, method = "complete")
plot(wisc.hclust)
abline(h=19, col="red", lty=2)
```


`Combining methods`
Clustering on PCA results

```{r}
wisc.pr.hclust <- hclust(data.dist, method = "ward.D2")
grps <- cutree(wisc.pr.hclust, k=2)
table(grps)
```

```{r}
table(grps, diagnosis)
```

```{r}
#plotting it
plot(wisc.pr$x[,1:2], col=grps)
```


```{r}
plot(wisc.pr$x[,1:2], col=diagnosis)
```

```{r}
g <- as.factor(grps)
levels(g)
```
```{r}
g <- relevel(g,2)
levels(g)
```

```{r}
# Plot using our re-ordered factor 
plot(wisc.pr$x[,1:2], col=g)
```
```{r}
library(rgl)
plot3d(wisc.pr$x[,1:3], xlab="PC 1", ylab="PC 2", zlab="PC 3", cex=1.5, size=1, type="s", col=grps)
```

```{r}
## Use the distance along the first 7 PCs for clustering i.e. wisc.pr$x[, 1:7]
dist <- dist(wisc.pr$x[,1:7])
wisc.pr.hclust <- hclust(dist, method="ward.D2")
```
```{r}
wisc.pr.hclust.clusters <- cutree(wisc.pr.hclust, k=2)
```

>Q15. How well does the newly created model with four clusters separate out the two diagnoses?

separates it out pretty well
```{r}
table(wisc.pr.hclust.clusters, diagnosis)
```

>Q16. How well do the k-means and hierarchical clustering models you created in previous sections (i.e. before PCA) do in terms of separating the diagnoses? Again, use the table() function to compare the output of each model (wisc.km$cluster and wisc.hclust.clusters) with the vector containing the actual diagnoses.

Hierarchical clustering is better since any number of clusters can be chosen that allow for better interpretation of the data better.
```{r}
wisc.km<-kmeans(scale(wisc.data), centers = 2, nstart = 20)
table(wisc.km$cluster, diagnosis)
table(wisc.hclust.clusters, diagnosis)
```

>Q17. Which of your analysis procedures resulted in a clustering model with the best specificity? How about sensitivity?

Hierarchical is best with specificity while K-means is best with sensitivity.


`Prediction`
Using predict() function on new cancer data

```{r}
#url <- "new_samples.csv"
url <- "https://tinyurl.com/new-samples-CSV"
new <- read.csv(url)
npc <- predict(wisc.pr, newdata=new)
npc
```

Plotting the new data
```{r}
plot(wisc.pr$x[,1:2], col=g)
points(npc[,1], npc[,2], col="blue", pch=16, cex=3)
text(npc[,1], npc[,2], c(1,2), col="white")
```

>Q18. Which of these new patients should we prioritize for follow up based on your results?

We should prioritize patient 1 since their cancer is malignant. 










