---
title: "Lab 17"
author: 'Thao Hoang PID: A15449860'
date: "11/23/2021"
output: html_document
---

```{r}
# Import vaccination data
vax <- read.csv("covid19vaccinesbyzipcode_test.csv")
head(vax)
```

> Q1. What column details the total number of people fully vaccinated?

persons_fully_vaccinated


> Q2. What column details the Zip code tabulation area? 

zip_code_tabulation_area

> Q3. What is the earliest date in this dataset? 

2021-01-05

> Q4. What is the latest date in this dataset?

```{r}
tail(vax)
```
2021-11-16


let’s call the skim() function from the skimr package to get a quick overview of this dataset:
```{r}
skimr::skim(vax)
```

> Q5. How many numeric columns are in this dataset? 

9

> Q6. Note that there are “missing values” in the dataset. How many NA values there in the persons_fully_vaccinated column? 

```{r}
sum( is.na(vax$persons_fully_vaccinated) )
```


> Q7. What percent of persons_fully_vaccinated values are missing (to 2 significant figures)? 
 
 1.2%
```{r}

sum( is.na(vax$persons_fully_vaccinated) )/sum(vax$persons_fully_vaccinated, na.rm=TRUE) *100
```


> Q8. [Optional]: Why might this data be missing?
Maybe people did not want to release their records due to privacy. 


##Working with dates

```{r}
library(lubridate)
today()
```

The as_of_date column of our data is currently not that usable. For example we can’t easily do math with it like answering the simple question how many days have passed since data was first recorded:
```{r}
# This will give an Error!
# today() - vax$as_of_date[1]
```

here we make our `as_of_date` column lubridate format
```{r}
# Speciffy that we are using the Year-mont-day format
vax$as_of_date <- ymd(vax$as_of_date)
```

Now I can do useful math with dates easily:
```{r}
#first entry
today() - vax$as_of_date[1]
```

> Q9. How many days have passed since the last update of the dataset?
How many days are between the first and last entry?

```{r}
#last entry
today() - vax$as_of_date[nrow(vax)]
```

```{r}
#how many days in between
vax$as_of_date[nrow(vax)] - vax$as_of_date[1]
```


> Q10. How many unique dates are in the dataset (i.e. how many different dates are detailed)?

```{r}
length(unique(vax$as_of_date))
```
```{r}
46*7
```

##Working with ZIP codes

```{r}
library(zipcodeR)
```

```{r}
geocode_zip('92037')
```

Calculate the distance between the centroids of any two ZIP codes in miles, e.g.
```{r}
zip_distance('92037','92109')
```

More usefully, we can pull census data about ZIP code areas (including median household income etc.). For example:
```{r}
reverse_zipcode(c('92037', "92109") )
```

We can use this reverse_zipcode() to pull census data later on for any or all ZIP code areas we might be interested in.
```{r}
# Pull data for all ZIP codes in the dataset
zipdata <- reverse_zipcode( vax$zip_code_tabulation_area )
```

##Focus on the San Diego area

```{r}
table(vax$county)

```

We will subset with base R
```{r}
inds <- vax$county == "San Diego"
head (vax[inds,])
```

But lets use **dplyr** package and it's **filter()** function

```{r}
library(dplyr)
sd <- filter(vax, county == "San Diego")
head(sd)
```
How many entries are there for the San Diego county?
```{r}
nrow(sd)
```
```{r}
sd.10 <- filter(vax, county == "San Diego" &
                age5_plus_population > 10000)
```


> Q11. How many distinct zip codes are listed for San Diego County? 

```{r}
length(unique(sd$zip_code_tabulation_area))
```

> Q12. What San Diego County Zip code area has the largest 12 + Population in this dataset?

```{r}
ind <- which.max(sd$age12_plus_population)
sd[ind,]
```
what is the population in the 92037 zip code area?
```{r}
filter(sd, zip_code_tabulation_area == "92037")[1,]
```

> Q13. What is the overall average “Percent of Population Fully Vaccinated” value for all San Diego “County” as of “2021-11-09”? 

```{r}
sd.now <- filter(sd, as_of_date == "2021-11-09")
head(sd.now)
mean(sd.now$percent_of_population_fully_vaccinated,na.rm = TRUE)
```

We can look at teh 6-number summary 
```{r}
summary(sd.now$percent_of_population_fully_vaccinated)
```

> Q14. Using either ggplot or base R graphics make a summary figure that shows the distribution of Percent of Population Fully Vaccinated values as of “2021-11-09”?
using base R plot

```{r}
hist(sd.now$percent_of_population_fully_vaccinated)
```
using ggplot
```{r}
library(ggplot2)

ggplot(sd.now) + aes(percent_of_population_fully_vaccinated) + geom_histogram(bins=15)
```

what about 92037 -UCSD/La Jolla

```{r}
filter(sd.now, zip_code_tabulation_area == "92037")

```
##Focus on UCSD/La Jolla

```{r}
ucsd <- filter(sd, zip_code_tabulation_area=="92037")
ucsd[1,]$age5_plus_population
```
using ggplot
> Q15. Using ggplot make a graph of the vaccination rate time course for the 92037 ZIP code area:

```{r}
ggplot(ucsd) +
  aes(as_of_date,percent_of_population_fully_vaccinated) +
  geom_point() +
  geom_line(group=1) +
  ylim(c(0,1)) +
  labs(x="date", y="Percent Vaccinated")
```
Comparing 92037 to other similar sized areas?

Let’s return to the full dataset and look across every zip code area with a population at least as large as that of 92037 on as_of_date “2021-11-16”.
```{r}
# Subset to all CA areas with a population as large as 92037
vax.36 <- filter(vax, age5_plus_population > 36144 &
                as_of_date == "2021-11-16")

head(vax.36)

```
> Q16. Calculate the mean “Percent of Population Fully Vaccinated” for ZIP code areas with a population as large as 92037 (La Jolla) as_of_date “2021-11-16”. Add this as a straight horizontal line to your plot from above with the geom_hline() function?

```{r}
mean(vax.36$percent_of_population_fully_vaccinated)
```

```{r}
ggplot(ucsd) +
  aes(as_of_date,percent_of_population_fully_vaccinated) +
  geom_point() +
  geom_line(group=1) + 
  geom_hline(yintercept = 0.663)+ 
  labs(x="date", y="Percent Vaccinated")
```

> Q17. What is the 6 number summary (Min, 1st Qu., Median, Mean, 3rd Qu., and Max) of the “Percent of Population Fully Vaccinated” values for ZIP code areas with a population as large as 92037 (La Jolla) as_of_date “2021-11-16”?

```{r}
summary(vax.36$percent_of_population_fully_vaccinated)
```


> Q18. Using ggplot generate a histogram of this data.

```{r}
ggplot(vax.36) +
  aes(percent_of_population_fully_vaccinated) +
  geom_histogram(bins=15)
```
>Q19. Is the 92109 and 92040 ZIP code areas above or below the average value you calculated for all these above?
92040 is below avg and 92109 is above avg. 

```{r}
vax %>% filter(as_of_date == "2021-11-16") %>%  
  filter(zip_code_tabulation_area=="92109") %>%
  select(percent_of_population_fully_vaccinated)
```


```{r}
vax %>% filter(as_of_date == "2021-11-16") %>%  
  filter(zip_code_tabulation_area=="92040") %>%
  select(percent_of_population_fully_vaccinated)
```





Population in the 92037 zip code area
```{r}
ucsd[1,]$age5_plus_population
```
First we need to subset the full `vax` dataset to include only the zip code areas with population as large as 92037

```{r}
vax.36.all <- filter(vax, age5_plus_population > 36144)
nrow(vax.36.all)
```

How many unique zip codes have a pop as large as 92037
```{r}
length(unique(vax.36.all$zip_code_tabulation_area))
```

>Q20. Finally make a time course plot of vaccination progress for all areas in the full dataset with a age5_plus_population > 36144.
Lets make a final figure that shows all these zip areas

```{r}
ggplot(vax.36.all)+ 
  aes(as_of_date,percent_of_population_fully_vaccinated, group=zip_code_tabulation_area) +
  geom_line(alpha=0.2) + 
  geom_hline(yintercept = 0.672, color ="red")+
  labs(x="date", y="Percent Vaccinated")
```
>Q21. How do you feel about traveling for Thanksgiving and meeting for in-person class next Week?
I don't really care what ended up happening. 






